---
title: "Component-wise Z-residual Diagnosis for Bayesian Hurdle Models"
output: 
  html_document:
    toc: true          # Enable table of contents
    toc_depth: 3       # Depth of headings in TOC
    toc_float: true    # TOC floats to the side
    number_sections: false
    theme: flatly      # Remove default Bootstrap theme
    includes:
      in_header: custom_header.html
bibliography: refs.bib       # Point to your BibTeX file
link-citations: yes
# runtime: shiny
date: "`r Sys.Date()`"
---

```{=html}
<style>
body, .container-fluid, .main-container {
  max-width: 100% !important;
  width: 95%;
  margin: auto;
  panmes.dataing: 20px;
}

h1 {
  color: darkred;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(animation)

```

```{r, echo=FALSE}
# Shortcuts for abbreviations
abbr <- function(short, full) {paste0('<abbr title="', full, '">', short, '</abbr>')}
RPP <- abbr("RPP", "Randomized Predictive P-value")
PMF <- abbr("PMF", "Probability Mass Function")
HNB <- abbr("HNB", "Hurdle Negative Binomial")
HP <- abbr("HP", "Hurdle Poisson")
SW <- abbr("SW", "Shapiro-Wilk")
ANOVA <- abbr("ANOVA", "Analysis of Variance")
BL <- abbr("BL", "Bartlett's")
```


## Introduction

This vignette demonstrates how to use the `Zresidual` package to compute component-wise Z-residuals for diagnosing Bayesian hurdle models [@Mullahy1986-em], based on the output from the `brms` package in R [@Burkner2017-brms] can be calculated separately for the zero, count and hurdle components to reveal potential model misspecifications. The examples illustrate the practical use of these residuals for `r RPP` [@Feng2020] diagnostics. For a detailed explanation of the methodology and theoretical foundations, please refer to the <mark>"To be updated"</mark>.

## Definitions of Component-wise Z-residuals for Bayesian Hurdle Models

This section demonstrates the definiotns of component-wise posterior predictive quantities including the posterior predictive `r PMF`, survival function, and `r RPP` for Bayesian hurdle models. Hurdle models consist of two parts:
<ul>
  <li>A **logistic component** modeling the probability of structural zeros.</li>
  <li>A **count component** modeling positive counts using a zero-truncated distribution.</li>
</ul>

Let $C_i \in \{0, 1\}$, where $C_i = 1$ indicates a non-zero value, and $C_i = 0$ indicates a zero value for the $i^\th$ observations. If $C_i=1$, the corresponding count model then operates on $y_i^+ \in \{1, 2, \dots\}$, i.e., the positive counts only.

Using Bayesian estimation (e.g., via the `brms` package), we draw $T$ samples from the posterior distribution. Let  $\theta^{(t)}$ denote the $t^\th$ posterior draw, including component-specific parameters like:
<ul>
  <li>$\hu$ : the zero porbability,</li>
  <li>$\mu_i^{(t)}, \phi^{(t)}$ : parameters for the count component.</li>
</ul>

For a given observation $y_i^\obs$, the component-wise posterior predictive `r PMF` and survival functions are defined below.

#### **Hurdle Model:**
\begin{equation}
p_i^{\post, \hurdle}(y_i^\obs) = \frac{1}{T} \sum_{t=1}^T
\begin{cases} 
{\hu}^{(t)} & \text{if } y_i^\obs = 0, \\
(1 - {\hu}^{(t)}) \cdot\frac{p_i^\ut(y_i^\obs | \theta^{(t)})}{1 - p_i^\ut(0 | \theta^{(t)})} & \text{if } y_i^\obs =1, 2, \ldots,\\
0 & \text{otherwise.}
\end{cases}  
\end{equation}
\begin{equation}
S_i^{\post, \hurdle}(y_i^\obs) = \frac{1}{T} \sum_{t=1}^T 
\begin{cases}
1 & \text{if } y_i^\obs < 0, \\
1-{\hu}^{(t)} & \text{if } 0 \le y_i^\obs < 1, \\
(1 - {\hu}^{(t)}) \cdot \frac{S_i^\ut(y_i^\obs \mid \theta^{(t)})}{1-p_i^\ut(0 \mid \theta^{(t)})} & \text{if } y_i^\obs \ge 1. 
\end{cases}
\end{equation}

#### **Logistic Component:**
\begin{equation}
\label{componentwise_logistic_pmf}
p_i^{\post, \logit}(c_i^\obs) = \frac{1}{T} \sum_{t=1}^T 
\begin{cases}
{\hu}^{(t)} & \text{if } c_i^\obs = 0, \\
1 - {\hu}^{(t)} & \text{if } c_i^\obs = 1,\\
0 & \text{otherwise.}
\end{cases}
\end{equation}

\begin{equation}
\label{componentwise_logistic_survival}
S_i^{\post, \logit}(c_i^\obs) = \frac{1}{T} \sum_{t=1}^T 
\begin{cases}
1 & \text{if } c_i^\obs < 0, \\
1-{\hu}^{(t)} & \text{if } 0 \le c_i^\obs < 1, \\
0, & \text{if } c_i^\obs \ge 1. 
\end{cases}
\end{equation}

#### **Count Compoenent:**
\begin{equation}
\label{componentwise_count_pmf}
p_i^{\post,\countm}({y_i^+}^\obs) = \frac{1}{T} \sum_{t=1}^T 
\begin{cases}
    \frac{p_i^\ut({y_i^+}^\obs \mid \theta^{(t)})}{1 - p_i^\ut(0 \mid \theta^{(t)})}, & \text{ for } {y_i^{+}}^\obs = 1,2,\ldots,\\
    0 & \text{ otherwise.}
\end{cases}
\end{equation}


\begin{equation}
\label{componentwise_count_survival}
S_i^{\post, \countm}({y_i^+}^\obs) = \frac{1}{T} \sum_{t=1}^T 
\begin{cases}
    1 & \text{ if } {y_i^+}^\obs < 1 \\
    \frac{S_i^\ut({y_i^+}^\obs \mid \theta^{(t)})}{1-p_i^\ut(0 \mid \theta^{(t)})}, & \text{ if } {y_i^+}^\obs \ge 1
\end{cases}
\end{equation}

where $p_i^\ut(. \mid \theta^{(t)})$ and $S_i^\ut(. \mid \theta^{(t)})$ denote the `r PMF` and survival function of the untruncated count distribution, given the component-specific posterior parameters $\theta^{(t)}$.

For any observed value $y_i^\obs$, we define:
\begin{equation}
\label{eq:post_rpp}\rpp_i(y_i^\obs | \theta^{(t)}) = S_i(y_i^\obs | \theta^{(t)}) + U_i \times p_i(y_i^\obs | \theta^{(t)})
\end{equation}
where $U_i \sim \uniform(0,1)$. Here, $y_i^\obs$ is the observed value, which can refer to either the binary response $C_i$ or the positive count ${y_i^+}^\obs$, depending on the component being evaluated. Then, the Z-residual of a discrete response variable is,
\begin{equation}
\label{eq:z_residual}
z_i = -\Phi^{-1}(\rpp_i(y_i^\obs|\theta)) \sim N(0, 1)
\end{equation}
where \(\Phi^{-1}(.)\) is the quantile function of standard normal distribution.

## Installation

<mark>To be updated<\mark>

## Examples for Illustration and Demonstration

### Dataset

This example illustrates the use of hurdle models to analyze healthcare utilization among older adults. The dataset is drawn from the National Medical Expenditure Survey (NMES), a nationally representative U.S. survey conducted in 1987–1988. A subsample of 4,406 individuals aged 66 and older, all enrolled in Medicare, is used. The primary response variable is the number of emergency department (ED) visits, with demographic, socioeconomic, and health-related covariates included in the analysis. 

```{r, include=FALSE}
# Not displayed in HTML
# All required packages
library(DT)
library(dplyr)

# PATH is set to Github
#path <- "https://github.com/dananjids/hurdlemodels/tree/main/"
path <- "C:/Users/ocf542/OneDrive - University of Saskatchewan/USask/Research/hurdle_model/demo/"
```

```{r, warning=FALSE}
library(hurdlemodels)
data("nmes.data")
```

```{r}
nmes.data$hlth <- NA  # Initialize with NA
nmes.data$hlth[nmes.data$exclhlth == 1] <- 2 # Set hlth = 2 for Excellent
nmes.data$hlth[nmes.data$poorhlth == 1] <- 0 # Set hlth = 0 for Poor
nmes.data$hlth[nmes.data$exclhlth == 0 & nmes.data$poorhlth == 0] <- 1 # Set hlth = 2 for Average
```

```{r echo = FALSE}
datatable(nmes.data,
              options = list(scrollX = TRUE, paging=TRUE), 
              caption = "Table 1: NMES dataset")
```

### Fit models

The data are modeled using hurdle Poisson (`r HP`) and hurdle negative binomial (`r HNB`) models via the `brms` package in R. These models allow to examine whether the logistic and count components offer additional, more localized insights compared to those obtained from the combined hurdle model.

```{r}
nmes.data1 <- nmes.data %>% select(c(emr, numchron, hlth, adldiff, school))
nmes.data2 <- nmes.data %>% select(c(emr, numchron, hlth, adldiff, school, black))
```

```{r}
predictors_hnb <- c("numchron", "hlth", "adldiff", "school")
predictors_hp <- c("numchron", "hlth", "adldiff", "school", "black")
count_formula_hnb <- formula(paste("emr ~ ", paste(predictors_hnb, collapse=" + ")))
zero_formula_hnb <- formula(paste("hu ~ ", paste(predictors_hnb, collapse=" + ")))
count_formula_hp <- formula(paste("emr ~ ", paste(predictors_hp, collapse=" + ")))
zero_formula_hp <- formula(paste("hu ~ ", paste(predictors_hp, collapse=" + ")))
```

```{r eval=FALSE}
fit_hnb <- brm(bf(count_formula_hnb, zero_formula_hnb),
               data = nmes.data1,
               family = hurdle_negbinomial(link = "log", link_shape = "identity", link_hu = "logit"),
               prior = c(prior(normal(0, 5), class = "b"),
                         prior(normal(0, 5), class = "Intercept"),
                         prior(gamma(0.01, 0.01), class = "shape"),
                         prior(normal(0, 2), class = "b", dpar = "hu")),
               refresh = 0)

fit_hp <- brm(bf(count_formula_hp, zero_formula_hp),
              data = nmes.data2,
              family = hurdle_negbinomial(link = "log",
                                          link_shape = "identity",
                                          link_hu = "logit"),
              prior = prior("normal(1000, 1)", class = "shape"), refresh = 0)
```

### Z-residual calculation

We computed Z-residuals for logistic, count and hurdle components separately using the Posterior `r RPP` method for the NMES dataset.

```{r eval=FALSE}
zres_hnb_zero <- zresidual_hurdle_negbinomial(fit_hnb, type = "zero", method = "post", nrep = 50)
zres_hnb_count <- zresidual_hurdle_negbinomial(fit_hnb, type = "count", method = "post", nrep = 50)
zres_hnb_hurdle <- zresidual_hurdle_negbinomial(fit_hnb, type = "hurdle", method = "post", nrep = 50)

zres_hp_zero <- zresidual_hurdle_poisson(fit_hp, type = "zero", method = "post", nrep = 50)
zres_hp_count <- zresidual_hurdle_poisson(fit_hp, type = "count", method = "post", nrep = 50)
zres_hp_hurdle <- zresidual_hurdle_poisson(fit_hp, type = "hurdle", method = "post", nrep = 50)
```

```{r eval=FALSE, echo=FALSE}
# ---Off comments when need to rerun the demo.html
# ---Not displayed in HTML
# ---Saving the loaded fits and zres Rdata to DEMO folder

# saveRDS(fit_hnb, file = paste0(path, "fit_hnb.Rds"))
# saveRDS(fit_hp, file = paste0(path, "fit_hp.Rds"))
# saveRDS(zres_hnb_zero, file = paste0(path, "zres_hnb_zero50.Rds"), compress = TRUE)
# saveRDS(zres_hnb_count, file = paste0(path, "zres_hnb_count50.Rds"), compress = TRUE)
# saveRDS(zres_hnb_hurdle, file = paste0(path, "zres_hnb_hurdle50.Rds"), compress = TRUE)
# saveRDS(zres_hp_zero, file = paste0(path, "zres_hp_zero50.Rds"), compress = TRUE)
# saveRDS(zres_hp_count, file = paste0(path, "zres_hp_count50.Rds"), compress = TRUE)
# saveRDS(zres_hp_hurdle, file = paste0(path, "zres_hp_hurdle50.Rds"), compress = TRUE)
```

```{r, echo=FALSE}
# ---Not displayed in HTML
# ---Loading the loaded fits and zres Rds from local disk
zres_hnb_zero <- readRDS(paste0(path, "zres_hnb_zero50.Rds"))
zres_hnb_count <-  readRDS(paste0(path, "zres_hnb_count50.Rds"))
zres_hnb_hurdle <-  readRDS(paste0(path, "zres_hnb_hurdle50.Rds"))
zres_hp_zero <-  readRDS(paste0(path, "zres_hp_zero50.Rds"))
zres_hp_count <-  readRDS(paste0(path, "zres_hp_count50.Rds"))
zres_hp_hurdle <-  readRDS(paste0(path, "zres_hp_hurdle50.Rds"))
```

### Graphically assess model

Figure 1 displays diagnostic plots of Z-residuals for the logistic, count, and overall components of the `r HNB` model. The residuals generally fall within the expected range (–3 to 3) without showing strong patterns, indicating a good model fit. Some outliers are present, particularly in the logistic component, as highlighted in the Q–Q plots. These outliers also appear in the overall model but are mostly driven by the logistic part. None of the residual-based statistical tests (`r SW`, `r ANOVA`, `r `BL`) yield significant p-values for any of the model components, providing no strong evidence of model misspecification. Collectively, these diagnostic results suggest that the hurdle negative binomial model offers an adequate fit to the data.

```{r  zresid_animation_hnb, fig.show='animate', interval=0.5, out.width='100%', animation.hook='gifski', results = 'hide', fig.cap="Figure 1: QQ Plots and Scatter plots of HNB for logistic, count and hurdle components.", fig.align='center'}
  for(i in 1:50){
    par(mfrow = c(2,3), mar = c(2, 2, 2, 2))
    qqnorm.zresid(zres_hnb_zero, irep = i, main.title = "HNB, Zero")
    qqnorm.zresid(zres_hnb_count, irep = i, main.title = "HNB, Count")
    qqnorm.zresid(zres_hnb_hurdle, irep = i, main.title = "HNB, Hurdle")

    plot.zresid(zres_hnb_zero, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
    plot.zresid(zres_hnb_count, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
    plot.zresid(zres_hnb_hurdle, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
  }
```

Figure 2 shows Z-residual diagnostics for the logistic, count, and hurdle components of the `r HP` model. The hurdle component exhibits greater variability and a funnel shape in the residuals versus fitted plot, along with large outliers and deviations from normality in the Q–Q plot, suggesting potential misspecification. In contrast, the logistic component displays more stable residuals and better alignment with theoretical quantiles. These observations are supported by the `r SW` test, which yields a significant p-value for the hurdle component and a non-significant result for the logistic component.

```{r  zresid_animation_hp, fig.show='animate', interval=0.5, out.width='100%', animation.hook='gifski', results = 'hide', fig.cap="QQ Plots and Scatter plots of HP for logistic, count and hurdle components.", fig.align='center'}
  for(i in 1:50){
    par(mfrow = c(2,3), mar = c(2, 2, 2, 2))
    qqnorm.zresid(zres_hp_zero, irep = i, main.title = "HP, Zero")
    qqnorm.zresid(zres_hp_count, irep = i, main.title = "HP, Count")
    qqnorm.zresid(zres_hp_hurdle, irep = i, main.title = "HP, Hurdle")

    plot.zresid(zres_hp_zero, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
    plot.zresid(zres_hp_count, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
    plot.zresid(zres_hp_hurdle, irep = i, X="fitted.value", outlier.return = T, main.title = NULL)
  }
```

## References
